---
title: "Module 4 Assignment Rmarkdown"
author: "Sarah Hodgson"
date: "2025-04-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 3.10 Assignment: End-to-end data analysis in R

This assignment is to develop a full end-to-end analysis of a dataset in R, designed to show how you can now take any dataset and use your skills as a marine data scientist to build impressive analyses and data visualisations that you can share with end-users. 

Note: this assignment relies on the skills developed in Workshop 3, including tidyr, to clean data.

In this assignment, I used real-world marine data collected as part of Queensland fisheries QFISH database to develop a report suitable for reporting to a manager, a community group or for my own research.

________________________________________________________________________________________________

### Part 1. Obtaining data. 

1. Review all of the available datasets in QFISH & identify which one I will analyse.

   Chosen dataset: Commercial Otter Trawl Fishery

2. Export the dataset in an appropriate format for R (CSV file).
3. Use my Github repo to save the data in an appropriate place.
Ensure I sync my github regularly by committing and pushing to the remote repository.

4. Read dataset & check if it displays correctly
```{r message=FALSE, warning=FALSE}
library(tidyverse)

#read csv file
qfish <- read.csv("../data/QFISH.csv")

qfish

```

### Part 2. Tidy the data.

5. Using the tidyverse, get my data ready for making a plot.

```{r}
# Clean the Tonnes column: remove commas and convert to numeric
qfish$Tonnes <- as.numeric(gsub(",", "", qfish$Tonnes))

```

6. Develop a plot to visualise the relationship between the number of fishing licences permitted and the tonnes of catch.

```{r message=FALSE, warning=FALSE}
# Remove the outlier (highest Tonnes value)
qfish_no_outlier <- qfish[qfish$Tonnes < max(qfish$Tonnes, na.rm = TRUE), ]

# Create scatter plot
licences_tonnes_plot <- ggplot() +
  geom_point(data = qfish_no_outlier, aes(x = Licences, y = Tonnes, color = "Catch Data"), # add data points, set x & y axes, colour Catch Data points
             alpha = 0.8, size = 1.5) + # transparency & size of data points
  geom_smooth(data = qfish_no_outlier, aes(x = Licences, y = Tonnes, color = "Trendline"), # add trend line, set x & y axes, colour Trendline
              method = "gam", se = FALSE, size = 1) + # use Generalised Additive Models (gam) for the trend line to make it fit the data more accurately
  scale_color_manual(
    name = "Legend", # make legend & name
    values = c("Catch Data" = "darkblue", "Trendline" = "red") # add colours to Catch data & Trend line
  ) +
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.15))) + # adjust scale
  labs(
    title = "Relationship Between Number of Licences and Tonnes of Catch", # name figure
    x = "Number of Licences", # name x axis
    y = "Tonnes of Catch" # name y axis
  ) +
  theme_minimal() # make theme

licences_tonnes_plot
```

7. Save plot & name, specify device to use, dimensions, units, & resolution.
```{r message=FALSE, warning=FALSE}
ggsave("licences_tonnes_plot.png", plot = licences_tonnes_plot, device = "png", width = 120, height = 80, units = "mm", dpi = 300)
```

