---
title: "MB5370 Module 04. Workshop 2 - Using ggplot2 for communication"
author: "Sarah Hodgson"
date: "2025-03-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

3.1 Labels

In ggplot2, you add labels with the labs() function. Let’s start with a title.

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(title = "Fuel efficiency generally decreases with engine size")

```

If you need to add more text, you can use a couple of other functions:
- subtitle adds additional detail in a smaller font beneath the title and caption adds text at the bottom right of the plot 
- caption adds text at the bottom right of the plot, often used to describe the source of the data.
```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(
    title = "Fuel efficiency generally decreases with engine size",
    subtitle = "Two seaters (sports cars) are an exception because of their light weight",
    caption = "Data from fueleconomy.gov"
  )

```

You can also use labs() to replace axis labels and legend titles.

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_smooth(se = FALSE) +
  labs(
    x = "Engine displacement (L)",
    y = "Highway fuel economy (mpg)",
    colour = "Car type"
  )

```

3.2 Annotations

But what if you want to add text to the plot itself? Sometimes you’ll want to do this to highlight a single observation, for instance.

Here you can use geom_text() to add textual labels to your plots. This works in the same way as geom_point() but rather than a shape geometry it can add a label. 

Here we will filter the data (more to come on this in the next workshop!) and add a label that calls in the values from the data frame. This is really neat, because it allows us to pull data values out (like the most fuel efficient car!) from the data itself and include it in the plot. This is particularly nice when these might change on an update of the data.

```{r}
#library(dplyr)
best_in_class <- mpg %>%
  group_by(class) %>%
  filter(row_number(desc(hwy)) == 1)

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_text(aes(label = model), data = best_in_class)

```

Although the labels overlap, there are ways you can handle that issue if you ever find it a problem. These include a nudge() function to move your text a certain amount and using other R packages to handle text wrapping etc. 

3.3 Scales

Changing the default scales on ggplot2 can help you customize your plots and improve communication of your results. 
Normally, ggplot2 automatically adds scales for you. For example, when you type:

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class))

```

ggplot2 automatically adds default scales behind the scenes.

However, if you want to tweak them, you can do so by offering values to the scale parameters by adding numbers in the appropriate scale arguments. 

Remember that to provide limits you’ll need to offer a character vector showing the start and end of your limit (e.g. c(0,12)). Note also that: ?scale_x_continuous() has plenty of other arguments, so don’t forget to explicitly state the argument you’re providing these limits for (e.g. limits = c(0,12)). 

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  scale_x_continuous() +
  scale_y_continuous() +
  scale_colour_discrete()

```

3.4 Axis ticks

You can also change the ticks on your axes. Breaks controls the position of the ticks and you can use labels as an argument to change the text label associated with ticks. 

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  scale_y_continuous(breaks = seq(15, 40, by = 5))

```

What does seq do? Try running 

```{r}
seq(15, 40, by = 5)
```

Similarly, you can use labels set to NULL to suppress the labels altogether. This is sometimes an option if you’re trying to format your plots in a particular way. 

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  scale_x_continuous(labels = NULL) +
  scale_y_continuous(labels = NULL)

```

3.5 Legends and colour schemes

Sometimes you might want to change the position of your legend, perhaps to make the plot itself as wide as possible (e.g. put the legend underneath) or to hide it all together. 
To control the overall position of the legend, you need to use a theme() setting. 
Run each line separately to see how it changes the plot of “base”.

```{r}
base <- ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class))

base + theme(legend.position = "left")
base + theme(legend.position = "top")
base + theme(legend.position = "bottom")
base + theme(legend.position = "right") # the default

```
You can also use legend.position = "none" to suppress the display of the legend altogether!

3.6	Replacing a scale

There are two types of scales you’re mostly likely to want to switch out: continuous position scales and colour scales.

It’s very useful to plot transformations of your variables. For example, as we’ve seen in diamond prices it’s easier to see the precise relationship between carat and price if we log transform them using a ggplot2 function for scale precisely for this purpose.

```{r}

ggplot(diamonds, aes(carat, price)) +
  geom_bin2d() + 
  scale_x_log10() + 
  scale_y_log10()

```

Another scale that is frequently customised is the colour scale. In fact, from a data visualisation point of view, this is among the most important changes you can make to a base ggplot.

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv))

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv)) +
  scale_colour_brewer(palette = "Set1")
```

If there are just a few colours, you can add a redundant shape mapping. This will also help ensure your plot is interpretable in black and white.

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv, shape = drv)) +
  scale_colour_brewer(palette = "Set1")
```

The ColorBrewer scales are documented online at http://colorbrewer2.org/. There are a range of different colour schemes you can import such as this, so read more about them in the ggplot2 cookbook: http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/ 

When you have predefined colours you want to use you can set them yourself, using scale_colour_manual()

```{r}
presidential %>%
  mutate(id = 33 + row_number()) %>%
  ggplot(aes(start, id, colour = party)) +
    geom_point() +
    geom_segment(aes(xend = end, yend = id)) +
    scale_colour_manual(values = c(Republican = "red", Democratic = "blue"))
```

Or one of the hottest colour packages out there now, even used in my own published work, the viridis colour scheme. Once you’ve tried this one you’ll start to see it everywhere!
To use it, simply use scale_colour_viridis() provided by the viridis package

```{r}
install.packages('viridis')
install.packages('hexbin')
library(viridis)
library(hexbin)

df <- tibble( # note we're just making a fake dataset so we can plot it
  x = rnorm(10000),
  y = rnorm(10000)
)
ggplot(df, aes(x, y)) +
  geom_hex() + # a new geom!
  coord_fixed()

ggplot(df, aes(x, y)) +
  geom_hex() +
  viridis::scale_fill_viridis() +
  coord_fixed()

```
You can even try some other sparky colour schemes, such as the Wes Anderson colour palette. Essential for those who love The Life Aquatic and other Wes Anderson films.


3.7 Themes

And finally, now you can customize the entire theme of your plot. Themes allow you to change some or all of the non-data elements of your plot with a theme.
ggplot2 has eight themes by default. Many more are included in add-on packages like ggthemes (https://github.com/jrnold/ggthemes), by Jeffrey Arnold.

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_bw()
```

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_light()
```

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_classic()
```

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_dark()
```

One particularly nice thing you can do is to develop your own themes simply by setting some or all of the arguments to theme() yourself. Here’s one I wrote so that my plots are always the same style:

```{r}
theme (panel.border = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position="bottom",
        legend.title=element_blank(),
        legend.text=element_text(size=8),
        panel.grid.major = element_blank(),
        legend.key = element_blank(),
        legend.background = element_blank(),
        axis.text.y=element_text(colour="black"),
        axis.text.x=element_text(colour="black"),
        text=element_text(family="Arial"))
```

3.8 Saving and exporting your plots
Now that you've iteratively developed your plot, you want to save it to disk in some format that allows you to import it into your report or to share it with the world. 
In general, the best way to make a single file will be to export it using the ggsave() function. 
There are two main ways to get your plots out of R using ggplot2. ggsave() will save the most recent plot to your working directory. 

```{r}
ggplot(mpg, aes(displ, hwy)) + geom_point()

ggsave("my-plot.pdf")
#> Saving 7 x 4.32 in image
```

Sometimes, you’ll want to play with the dimensions of your plot, in which case you can use the width and height arguments to this function. Try it!
Note that you can alternatively export your results, together with your code, using R markdown via knitr. 

** If you are up to this point it is time to work on your Plot Deconstruction exercise. Please continue with this for the remainder of this workshop.**

_________________________________________________________________________________________

3.10 Assignment: End-to-end data analysis in R

Once you’ve completed your first assignment for this module (Plot Deconstruction), you can move on to developing a full end-to-end analysis of a dataset in R. This task is designed to show you how you can now take any dataset and use your skills as a marine data scientist to build impressive analyses and data visualisations that you can share with end-users. 

Note: this assignment will be introduced in detail in Workshop 3, because it relies on the tidyr skills that you’ll develop in the next workshops for this module.


